A new model to predict survival in patients with end-stage liver disease

### 2022-07-22 analysis/article.Rmd
refactor: apply proofreading suggestions

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A new model to predict survival in patients with end-stage liver disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage in grafts for liver transplantation requires
      risk stratification and adequate allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD)
      score for 90-day mortality prediction with the help of different
      machine learning (ML) algorithms.

  **Methods**:
      We retrospectively analyzed clinical and laboratory data of 654
      patients who were recruited during the evaluation process for
      liver transplantation at the University Hospital Leipzig.
      After comparing 13 different ML algorithms in a nested cross-validation
      setting and selecting the best performing one we build a new model
      to predict 90-day mortality in patients with end-stage liver disease.

  **Results**:
      Penalized regression algorithms yield the highest prediction performance
      in our ML algorithm benchmark.
      In favour of a simpler model we chose the least absolute shrinkage and
      selection operator (lasso) regression.
      Beside the classical MELD variables INR and bilirubin the
      lasso regression selected cystatin C over creatinine, and
      additionally IL-6, total protein and cholinesterase.
      The new model offers an improved discrimination and calibration than MELD
      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new ML based model of end-stage liver disease that
      outperforms the classical MELD score for 90-day survival prediction.

  **Keywords**:
      end-stage liver disease; IL-6; machine learning; MELD;
      risk stratification; survival estimation.

@@

Liver cirrhosis is the terminal result of fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible and
the only available therapy is liver transplantation.
However, the shortage in grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates the patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine and international normalized
ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) as this was found to be an important additional risk factor in
liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed for predicting the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
Afterwards it was revalidated for predicting mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score especially the
creatinine and INR are highly dependent on the utilized laboratory methods
[@trotter2004; @cholongitas2007].
Additionally women are disadvantage [@sealock2022].
Patients with identical disease state could have very different MELD scores and
thus get different priority on the liver transplantation waiting list.
Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
underestimates the mortality risk [@hernaez2020].

There have been some attempts to use the data extracted from more than 300.000
electronic medical records from two hospitals in the United States to improve
the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
add albumin, white blood cell count, total cholesterol, age and length of
stay to the MELD-Na variables. Despite its published prediction improvement the
MELD-Plus scores are not used for transplant allocation yet.

As depicted by MELD-Plus risk scores better predictive scores often need more
variables and are more complicated. To reduce the risk of overlooking or
incorrectly calculating and interpreting the results,
clinical decision support systems may be used and could improve patient safety.
The research project on digital laboratory medicine (AMPEL) develops a
clinical decision support system based on laboratory diagnostics that should
support clinical practitioners in interpreting the laboratory results and taking
the necessary medical interventions [@eckelt2020].

This study aims to find clinical and laboratory values that improve the risk
stratification for liver transplantation over classical MELD, MELD-Na and
MELD-Plus scores and could be implemented as part of the AMPEL clinical decision
support system.

@@

In a retrospective cohort study we followed `r n_total` consecutive
patients, who were recruited during the evaluation process for
liver transplantation at the University Hospital of Leipzig from
`r paste(format(attr(raw_data, "period"), "%B %Y"), collapse = " to ")`.
For each patient we recorded `r ncol(raw_data)` variables.
Among them are age, sex, etiology of liver disease, complications as listed in
Table \@ref(tab:tbl1) and `r length(cn_labs)` laboratory measurements.

```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart Inclusion/Exclusion."}
library("consort")
library("grid")

@@

total_records <- paste0("Total records (n = ", n_total, ")")
analysed_records <- paste0("Analysed records (n = ", n_analysed, ")")
excluded_records <- paste0(
    "Excluded (n = ", sum(n_excluded), ")\n",
    paste0(
        "- ",
        c("Younger than 18 years",
          "Liver transplantation before visit",
          "Follow-up missing/invalid"
        ),
        " (n = ", n_excluded, ")",
        collapse = "\n"
    )
)

@@

We excluded `r sum(n_excluded)` patients from our analysis
(Figure \@ref(fig:flowchart)).
`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years.
`r n_excluded["LiverTransplantation"]` had a liver transplantation before and
in `r n_excluded["LostFollowUp"]` cases the follow up data were missing or
invalid.

@@

MELD and MELD-Na were calculated as described in
@optn_policies2021 using the following formulas:
$MELD = 10 * (0.957 \ln(creatinine [mg/dL]) + 0.378 \ln(bilirubin [mg/dL]) + 1.120 * \ln(INR) + 0.643)$,
Creatinine, bilirubin and INR values lower than 1.0 mg/dL were set to 1.0 mg/dL.
The maximum allowed creatinine was 4.0 mg/dL. If patients received dialysis,
creatinine was set to 4.0 mg/dL.
For MELD-Na MELD was calculated as above and recalculated if greater than 11 using:
$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/L]) - (0.033 * MELD_i * (137 - Na [mmol/L]))$

@@

The MELD-Plus9 score was ignored because the length of stay information was not
available for our cohort and the model itself wasn't superior to MELD-Plus7 in
its original publication [@kartoun2017].

@@

All data processing, statistical and machine learning analyses were performed
using R version
`r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
[@R-base].
Prior to the analyses all laboratory values were *zlog* transformed to
approximate a normal distribution as described in
@hoffmann2017 and implemented in @R-zlog.

@@

Where $LL$ and $UL$ are the lower and upper reference limit of
the corresponding laboratory diagnostic, respectively.
Missing laboratory measurements were replaced with a *zlog* value of zero.
In similar manner if information about complications or
comorbidities were missing we treat them as not present.

We compared `r n_lrns` different statistical and machine learning
algorithms using the *mlr3* framework
[@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
The used algorithms are designed for analysis of survival data:
Cox proportional hazards regression model
[@cox1972; @survival-book; @R-survival],
penalized regressions,
namely two different implementations of the
lasso (least absolute shrinkage and selection operator) regression
[@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
two different implementations of the
ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
the elastic net regression [@glmnet2011; @R-glmnet],
two different random forest implementations
[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC],
extreme gradient boosting [@chen2016; @R-xgboost],
two different support vector machines [@vanbelle2011; @R-survivalsvm]
and two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].

We used nested resampling to avoid feature- and
model-selection-bias to get reliable metrics to choose the best model.
Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
and a `r crossval$outer$param_set$values$folds`-fold outer cross-validation
that was repeated `r crossval$outer$param_set$values$repeats` times.
We searched the hyperparameter space using a grid search with a resolution of
`r tuner$param_set$get_values()$resolution`.
The models were evaluated using the concordance index, which is a measure of
rank correlation between predicted and observed events [@harrell1982].
Subsequently we selected the model with the highest median concordance index.
If multiple models were comparable we chose the simpler (more penalized) one.

After selecting the final model we trained and validated it as described in
@harrell1996.
Briefly we developed our model using a `r ameldcfg$nfold` cross-validation
that was repeated `r ameldcfg$nrepcv` times and finally predicted survival
at `r ameldcfg$times` days.
The predictions were sorted and grouped into equal sized intervals so that there
are `r ameldcfg$m` patients in each interval.
For each group the mean survival probability and the difference to the
Kaplan-Meier estimate was calculated.
The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
Subsequently the differences between predicted survival and Kaplan-Meier
estimates of the bootstrap samples were averaged and added to the original
model difference to obtain a bias-corrected estimated between predicted and
observed survival.

The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
using the area under the time-dependent receiver operating characteristic curve
(AUROC) based on the nonparametric inverse probability of censoring weighting
estimator (IPCW) [@blanche2013, @R-timeROC].
Testing was applied as described in @blanche2013.
A $p < 0.05$ was considered as statistical significant difference.

@@

In this study we analyzed a cohort of `r n_analysed` patients evaluated for
liver transplantation (Tab. \@ref(tab:tbl1)).
`r sprintf("%i (%s)", sum(raw_data$Sex == "male"), gtsummary::style_percent(mean(raw_data$Sex == "male"), symbol = TRUE))`
of them were men.
The median (IQR) follow-up times was `r .tbl1("DaysAtRisk")` days.
Within the follow-up time `r .tbl1("LTx")` patients received a liver transplant
and were censored for the analysis beyond this time point.
The median (IQR) age of the cohort was `r .tbl1("Age")` years.

Most of the patients presented with cirrhosis
(`r .tbl1("Cirrhosis")`).
The remaining patients had an acute liver failure
(`r .tbl1("ALF")`) or
a chronic liver insufficiency
(`r .tbl1("CLI")`).
At evaluation time the median (IQR) MELD and MELD-Na score were
`r .tbl1("ScoreMeldUnos")` and
`r .tbl1("ScoreMeldNaUnos")`, respectively.
Within 90 days `r .tbl1("M90")` patients died.
In nearly two third of all patients alcohol was the main cause
(`r .tbl1("Ethyltoxic")`) of their end-stage liver disease.
The remaining patients had viral infections
(hepatitis B and C viruse `r .tbl1("HBV")` and `r .tbl1("HCV")`, respectively),
autoimmune processes (autoimmune hepatitis `r .tbl1("AIH")`,
primary sclerosing cirrhosis `r .tbl1("PSC")`),
primary biliary cirrhosis `r .tbl1("PBC")` or unknown causes
(non-alcoholic steatohepatitis `r .tbl1("NASH")` and
cryptogenic hepatitis `r .tbl1("Cryptogenic")`).
The most common complications among all patients were gastrointestinal
bleedings in `r .tbl1("GIB")` patients, followed by
hepatocellular carcinoma (HCC) in `r .tbl1("HCC")` and
spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
Just `r .tbl1("Dialysis")` patients required a renal dialysis.

@@

According to @wiesner2003 we divided our cohort into
`r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
respectively.
As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality)
our observed 90-day mortality was
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times higher than the predicted one. Just in the group with the lowest MELD
scores the mortality was much lower.

@@

The penalized regression and random forest models
result in a similar high median performance in the benchmark process.
In favour of simpler models we chose a lasso based approach for the final model
development.
The classical cox model, the data and computational intense algorithms
extreme gradient boosting, support vector machines and neural networks yield a
worse performance.

@@

The resulting model AMPEL model of end-stage liver disease
(AMELD) uses `r length(cfs)` predictors.
Beside the classical MELD variables INR and (direct) bilirubin the lasso
regression selected cystatin C over creatinine, and
additionally IL-6, total protein and cholinesterase.

@@

The 90-day survival estimate of a given patient could be calculated as shown in
\@ref(eq:survcalc).

@@

Where $S_{0}$ is the baseline survival estimate from
\@ref(tab:basehaz) and $LP$ is the linear predictor of the coefficients
in \@ref(tab:coeftbl).

@@

```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
col <- viridisLite::viridis(5)
m <- list(
    AMELD = timeROC_RCV,
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD-Plus7" = timeROC_MELDPlus7
)
invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
roc <- roc[order(-roc)]
```

@@

AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
However, after correction for multiple testing the difference against
the second best model `r names(roc)[2]`
(AUROC: `r round(roc[2], 3)`) is not statistical significant
(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
(see Table \@ref(tab:roctest) for complete comparison).

```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival. The error bars correspond to the 95 % confidence interval of the survival estimate.'), echo = FALSE}
par(cex = 0.9)
plot(
    bootrcv,
    what = "calibration",
    main = "90-day survival",
    legend = FALSE
)

@@

Fig. \@ref(fig:bootrcv) depicts the calibration curve for AMELD and the
competing MELD scores.
While the prediction for low-risk patients is similar to other scores
AMELD predicts the medium-risk patients more accurate.
Because of the low number of very ill patients in our cohort the prediction
performance in high-risk patients is poor but comparable to the other models.
This would require a recalibration in a larger validation cohort.

@@

In this retrospective analysis of `r n_analysed` consecutive patients,
who were recruited during the evaluation process for liver transplantation at
the University Hospital of Leipzig we developed a new model to predict the
90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
score and the recently published MELD-Plus7 risk score.

@@

rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
rfranktxt <-
    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
```
Although the Organ Procurement and Transplantation Network recommends the use
of MELD-Na over MELD the sodium does not add any information to the score.
Neither in the bootstrap selection nor in the variable importance of a
random forest model the sodium gains any important weight
(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
In addition sodium levels can be affected by diuretic therapy and already
@kartoun2017 have shown that the discrimination performance of MELD-Na is
insignificantly different from the classical MELD.
Consequentially it is not a component of our model.

It has shown that women receive in median lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@sealock2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017] in our study.
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods.
It also reduce the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

Our model, AMELD, uses `r length(cfs)` predictors.
Half of them, INR, (direct) bilirubin, and cystatin C are identical
or similar to the labortory measurements utilized in the classical MELD.
The direct bilirubin selected by our approach and the total bilirubin in MELD
are highly correlated (spearman correlation:
`r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"], 2)`) and
therefore interchangeable.

In our feature selection process cystatin C was chosen instead of creatinine.
This is inline with recent studies that show a better prediction of mortality
for cystatin C in unselected and critical ill patients
[@helmerssonkarlqvist2016; @helmerssonkarlqvist2021]

Furthermore AMELD extends MELD by additional synthesis parameters
total protein and cholinesterase and
most importantly the inflammatory marker IL-6.

The latter is an example where results from machine learning can aid in
finding clinical meaningful patterns and guide further research.
The systemic inflammatory response syndrome is an
independent risk factor for poor outcome in patients with cirrhosis
[@thabut2007].
However neither MELD nor MELD-Na taking inflammation into account.
In contrast IL-6 is the most important variable in our model. Even if IL-6 is
excluded the algorithms select CRP as one of the most important variables which
highlights the important role of inflammation in the prognosis of patients
with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increase the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

>>>>

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A new model to predict survival in patients with end-stage liver disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage of grafts for liver transplantation requires
      risk stratification and adequate allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD)
      score for 90-day mortality prediction with the help of different
      machine-learning algorithms.

  **Methods**:
      We retrospectively analyzed the clinical and laboratory data of 654
      patients who were recruited during the evaluation process for
      liver transplantation at University Hospital Leipzig.
      After comparing 13 different machine-learning algorithms in a nested
      cross-validation setting and selecting the best performing one we build
      a new model to predict 90-day mortality in patients with end-stage liver
      disease.

  **Results**:
      Penalized regression algorithms yielded the highest prediction
      performance in our machine-learning algorithm benchmark.
      In favor of a simpler model we chose the least absolute shrinkage and
      selection operator (lasso) regression.
      Beside the classical MELD variables international normalized ratio (INR)
      and bilirubin, the lasso regression selected cystatin C over creatinine,
      as well IL-6, total protein, and cholinesterase.
      The new model offers improved discrimination and calibration over MELD
      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new machine-learning-based model of end-stage liver disease
      that outperforms the classical MELD score for 90-day survival prediction.

  **Keywords**:
      end-stage liver disease; IL-6; machine learning; MELD;
      risk stratification; survival estimation

@@

Liver cirrhosis is the terminal result of the fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible, and
the only available therapy is liver transplantation.
However, the shortage of grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine, and
the international normalized ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) because this was found to be an important additional
risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed to predict the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
It was subsequently revalidated to predict mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score,
creatinine and INR are highly dependent on the laboratory methods used
[@trotter2004; @cholongitas2007].
In addition women are disadvantaged [@sealock2022].
Patients with identical disease states can have very different MELD scores and
thus receive different priority levels on the liver transplantation waiting
list.
Furthermore, for acute-on-chronic liver failure, for example, the MELD score
often underestimates the mortality risk [@hernaez2020].

There have been some attempts to use the data extracted from more than 300,000
electronic medical records from two hospitals in the United States to improve
the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9
risk scores add albumin, white blood cell count, total cholesterol, age, and
length of stay to the MELD-Na variables.
Despite their published prediction improvement, the MELD-Plus scores are not
yet used for transplant allocation.

As shown by the MELD-Plus risk scores, better predictive scores often need more
variables and are more complicated. To reduce the risk of overlooking or
incorrectly calculating and interpreting the results,
clinical decision support systems may be used and could improve patient safety.
The research project on digital laboratory medicine (AMPEL) develops a
clinical decision support system based on laboratory diagnostics that should
support clinical practitioners in interpreting laboratory results and taking
the necessary medical interventions [@eckelt2020].

This study aims to find clinical and laboratory values that improve on the risk
stratification for liver transplantation of the classical MELD, MELD-Na, and
MELD-Plus scores and that might be implemented as part of the
AMPEL clinical decision support system.

@@

In a retrospective cohort study we followed `r n_total` consecutive
patients, who were recruited during the evaluation process for
liver transplantation at University Hospital Leipzig from
`r paste(format(attr(raw_data, "period"), "%B %Y"), collapse = " to ")`.
For each patient, we recorded `r ncol(raw_data)` variables, including
age, sex, etiology of liver disease, complications as listed in
Table \@ref(tab:tbl1), and `r length(cn_labs)` laboratory measurements.

```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart showing inclusion and exclusion of records."}
library("consort")
library("grid")

@@

total_records <- paste0("Total records (n = ", n_total, ")")
analysed_records <- paste0("Analyzed records (n = ", n_analysed, ")")
excluded_records <- paste0(
    "Excluded (n = ", sum(n_excluded), ")\n",
    paste0(
        "- ",
        c("Younger than 18 years",
          "Liver transplantation before visit",
          "Follow-up missing/invalid"
        ),
        " (n = ", n_excluded, ")",
        collapse = "\n"
    )
)

@@

We excluded `r sum(n_excluded)` patients from our analysis
(Figure \@ref(fig:flowchart)).
`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years,
`r n_excluded["LiverTransplantation"]` had had a liver transplant before and
in `r n_excluded["LostFollowUp"]` cases the follow-up data were missing or
invalid.

@@

MELD and MELD-Na were calculated as described in
@optn_policies2021 using the following formulas:
$MELD = 10 * (0.957 \ln(creatinine [mg/dL]) + 0.378 \ln(bilirubin [mg/dL]) + 1.120 * \ln(INR) + 0.643)$,
Creatinine, bilirubin and INR values lower than 1.0 mg/dL were set to 1.0 mg/dL.
The maximum allowed creatinine was 4.0 mg/dL. If the patients received dialysis,
creatinine was set to 4.0 mg/dL.
For MELD-Na, MELD was calculated as above and recalculated if greater than
11 using:
$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/L]) - (0.033 * MELD_i * (137 - Na [mmol/L]))$

@@

The MELD-Plus9 score was ignored because the length of stay information was not
available for our cohort, and the model itself was not superior to MELD-Plus7
in its original publication [@kartoun2017].

@@

All data processing, statistical, and machine-learning analyses were performed
using R version
`r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
[@R-base].
Prior to the analyses, all laboratory values were *zlog* transformed to
approximate a normal distribution as described in
@hoffmann2017 and implemented in @R-zlog.

@@

where $LL$ and $UL$ are the lower and upper age-, and sex-related
reference limits of the corresponding laboratory diagnostic, respectively.
Missing laboratory measurements were replaced with a *zlog* value of zero.
In a similar manner, where information about complications or
comorbidities was missing, we treated them as not present.

We compared `r n_lrns` different statistical and machine-learning
algorithms using the *mlr3* framework
[@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
The algorithms used were designed for analysis of survival data:
Cox proportional hazards regression model
[@cox1972; @survival-book; @R-survival];
penalized regressions,
namely, two different implementations of the
lasso (least absolute shrinkage and selection operator) regression
[@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
two different implementations of the
ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
the elastic net regression [@glmnet2011; @R-glmnet];
two different random forest implementations
[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC];
extreme gradient boosting [@chen2016; @R-xgboost];
two different support vector machines [@vanbelle2011; @R-survivalsvm];
;nd two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].

We used nested resampling to avoid feature- and
model-selection-bias to obtain reliable metrics to choose the best model.
We thus applied a `r crossval$inner$param_set$values$folds`-fold inner
and `r crossval$outer$param_set$values$folds`-fold outer cross-validation
that was repeated `r crossval$outer$param_set$values$repeats` times.
We searched the hyperparameter space using a grid search with a resolution of
`r tuner$param_set$get_values()$resolution`.
The models were evaluated using the concordance index, which is a measure of
rank correlation between predicted and observed events [@harrell1982].
Subsequently, we selected the model with the highest median concordance index.
If multiple models were comparable, we chose the simpler (more penalized) one.

After selecting the final model, we trained and validated it as described in
@harrell1996.
Briefly, we developed our model using a `r ameldcfg$nfold`-fold
cross-validation that was repeated `r ameldcfg$nrepcv` times and finally
predicted survival at `r ameldcfg$times` days.
The predictions were sorted and grouped into equal-sized intervals so that there
were `r ameldcfg$m` patients in each interval.
For each group, the mean survival probability and the difference to the
Kaplan-Meier estimate was calculated.
The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
Subsequently, the differences between predicted survival and Kaplan-Meier
estimates of the bootstrap samples were averaged and added to the original
model difference to obtain a bias-corrected estimate between predicted and
observed survival.

The final model was compared against the MELD, MELD-Na, and MELD-Plus7 scores
using the area under the time-dependent receiver operating characteristic curve
(AUROC) based on the nonparametric inverse probability of censoring weighting
estimator (IPCW) [@blanche2013, @R-timeROC].
Testing was applied as described in @blanche2013.
A $p < 0.05$ was considered a statistically significant difference.

@@

In this study, we analyzed a cohort of `r n_analysed` patients evaluated for
liver transplantation (Tab. \@ref(tab:tbl1)).
Of these,
`r sprintf("%i (%s)", sum(raw_data$Sex == "male"), gtsummary::style_percent(mean(raw_data$Sex == "male"), symbol = TRUE))`
were men.
The median (IQR) follow-up times was `r .tbl1("DaysAtRisk")` days.
Within the follow-up time, `r .tbl1("LTx")` patients received a liver transplant
and were censored for the analysis beyond this time point.
The median (IQR) age of the cohort was `r .tbl1("Age")` years.

Most of the patients presented with cirrhosis
(`r .tbl1("Cirrhosis")`).
The remaining patients had acute liver failure
(`r .tbl1("ALF")`) or
chronic liver insufficiency
(`r .tbl1("CLI")`).
At evaluation time, the median (IQR) MELD and MELD-Na scores were
`r .tbl1("ScoreMeldUnos")` and
`r .tbl1("ScoreMeldNaUnos")`, respectively.
Within 90 days, `r .tbl1("M90")` patients had died.
In nearly two thirds of all patients
(`r .tbl1("Ethyltoxic")`), alcohol was the main cause
of their end-stage liver disease.
The remaining patients had viral infections -
hepatitis B (`r .tbl1("HBV")`) or C (`r .tbl1("HCV")`) -
autoimmune autoimmune (`r .tbl1("AIH")`),
primary sclerosing cirrhosis (`r .tbl1("PSC")`),
primary biliary cirrhosis (`r .tbl1("PBC")`), unknown causes
(non-alcoholic steatohepatitis (`r .tbl1("NASH")`), or
cryptogenic hepatitis (`r .tbl1("Cryptogenic")`)).
The most common complications among all patients were gastrointestinal
bleedings in `r .tbl1("GIB")` patients, followed by
hepatocellular carcinoma (HCC) in `r .tbl1("HCC")` and
spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
Just `r .tbl1("Dialysis")` patients required renal dialysis.

@@

According to @wiesner2003, we divided our cohort into
`r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
respectively.
As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality),
our observed 90-day mortality was
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times higher than the predicted one.
Only in the group with the lowest MELD scores was the mortality much lower.

@@

The penalized regression and random forest models
resulted in a similar high median performance in the benchmark process.
In favour of simpler models, we chose a lasso-based approach for the final model
development.
The classical Cox model, the data and computational intense algorithms
extreme gradient boosting, the support vector machines and the neural networks
yielded a worse performance.

@@

The resulting AMPEL model of end-stage liver disease
(AMELD) uses `r length(cfs)` predictors.
Besides the classical MELD variables INR and (direct) bilirubin,
the lasso regression selected cystatin C over creatinine, as well as
IL-6, total protein, and cholinesterase.

@@

The 90-day survival estimate of a given patient can be calculated as shown in
\@ref(eq:survcalc).

@@

where $S_{0}$ is the baseline survival estimate from
\@ref(tab:basehaz) and $LP$ is the linear predictor of the coefficients
in \@ref(tab:coeftbl).

@@

```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve. Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
col <- viridisLite::viridis(5)
m <- list(
    AMELD = timeROC_RCV,
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD-Plus7" = timeROC_MELDPlus7
)
invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
roc <- roc[order(-roc)]
```

@@

AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) of all competitors.
However, after correction for multiple testing, the difference against
the second-best model, `r names(roc)[2]`
(AUROC: `r round(roc[2], 3)`), is not statistical significant
(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
(see Table \@ref(tab:roctest) for complete comparison).

```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival. The error bars correspond to the 95% confidence interval of the survival estimate.'), echo = FALSE}
par(cex = 0.9)
plot(
    bootrcv,
    what = "calibration",
    main = "90-day survival",
    legend = FALSE
)

@@

Fig. \@ref(fig:bootrcv) depicts the calibration curve for AMELD and the
competing MELD scores.
While the prediction for low-risk patients is similar to other scores,
AMELD predicts the medium-risk patients more accurately.
Because of the low number of very ill patients in our cohort, the prediction
performance in high-risk patients is poor but comparable to the other models.
This would require recalibration in a larger validation cohort.

@@

In this retrospective analysis of `r n_analysed` consecutive patients
who were recruited during the evaluation process for liver transplantation at
University Hospital Leipzig, we developed a new model to predict
90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
score and the recently published MELD-Plus7 risk score.

@@

rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
rfranktxt <-
    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
```

Although the Organ Procurement and Transplantation Network recommends the use
of MELD-Na over MELD, the sodium does not add any information to the score.
Neither in the bootstrap selection nor in the variable importance of a
random forest model does the sodium gain any important weight
(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
In addition, sodium levels can be affected by diuretic therapy, and
the discrimination performance of MELD-Na has already been shown by
@kartoun2017 to be insignificantly different from the classical MELD.
Consequently it is not a component of our model.

It has been shown that women receive median lower scores in MELD-Na and
therefore are less likely to receive an urgent listing for liver transplantation
[@sealock2022].
To address sexual differences, in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017] in our study.
This transformation also yields comparable laboratory measurements
across different clinical laboratories and methods, as well as reducing
the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample-dependent scaling prior to model development.
Both should increase the external applicability of our approach.

Our model, AMELD, uses `r length(cfs)` predictors.
Three of these - INR, (direct) bilirubin, and cystatin C - are identical
or similar to the laboratory measurements utilized in the classical MELD.
The direct bilirubin selected by our approach and the total bilirubin in MELD
are highly correlated (Spearman correlation:
`r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"], 2)`) and
therefore interchangeable.

In our feature selection process, cystatin C was chosen instead of creatinine.
This is in line with recent studies that show a better prediction of mortality
for cystatin C in unselected and critically ill patients
[@helmerssonkarlqvist2016; @helmerssonkarlqvist2021].

Furthermore AMELD, extends MELD by the additional synthesis parameters
total protein and cholinesterase, and
most importantly the inflammatory marker IL-6.

The latter is an example where results from machine-learning can aid in
finding clinically meaningful patterns and guide further research.
Systemic inflammatory response syndrome is an
independent risk factor for poor outcomes in patients with cirrhosis
[@thabut2007].
However, neither MELD nor MELD-Na took inflammation into account.
By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
excluded, the algorithms select CRP as one of the most important variables,
which highlights the important role of inflammation in the prognosis of
patients with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].
